---
title: "Linear Modeling"
author: "Haley Reed"
date: "6/8/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

To better understand polls, we can attempt to model the bias from polls with other measurable variables. 

```{r}
#load the data
#in this case from 538
mydata <- read.csv("raw-polls_538.csv")
```

Choose what years you would like to consider:

```{r, echo=TRUE}
#Establish which years you wish to use data from (between 1998 and 2020):
earlyYear=1998
lateYear=2020
```

Choose the minimum number (at least 50) of polls you wish a pollster to have before they are considered in the model:

```{r, echo=TRUE}
minPolls=100
```

```{r}
#subset to just even years to avoid special elections - harder to model
mydata = mydata[mydata$year %% 2 == 0, ]
#subset to preferred years
mydata=(subset(mydata, mydata$year>=earlyYear))
mydata=(subset(mydata, mydata$year<=lateYear))


#use lubridate to change dates from character to date data type for functionality. 
mydata$electiondate = lubridate::mdy(mydata$electiondate)
mydata$polldate = lubridate::mdy(mydata$polldate)

#subset to make sure it is a democratic v. republican race
mydata=subset(mydata,cand1_party=="DEM")
mydata=subset(mydata,cand2_party=="REP")

mydata$year = as.factor(mydata$year)

#create empty data frame, while maintaining all columns from the mydata structure
myDataMyPollsters=  mydata[0,]



#run a loop to fill this data frame with every pollster
for(myPollster in unique(mydata$pollster))
{
  #subset to a dataset with just each pollster
  subpoll=subset(mydata, pollster==myPollster)
  #if that subset has more than thirty entries, use it:
  if(nrow(subpoll)>minPolls){
  #Use only the most recent poll for each election:
  #Use setDT function from data.table package to get a subset from mydata with just the max. value of the date element for each race (grouped with the keyby function). Call this new subset of data onlyRecentData.
  onlyRecentData1=(data.table::setDT(subpoll)[,.SD[which.max(polldate)],keyby=race_id])
  #combine each of these subsets together
  myDataMyPollsters=rbind(myDataMyPollsters, onlyRecentData1)
  }
}


#myDataMyPollsters = filter(myDataMyPollsters, ((year %% 2) == 0) )

```

### Data Exploration


Before modeling our data, we have a few variables in mind. Of consideration are:

* The type of race: senate, house, presidential?

* The year of the election 

* The pollster conducting the poll

* Is the poll partisan? If so, which party funded it?

* What is the sample size of the poll?

It is worth separating out between pollsters. A density plot shows this:
```{r}
#density plot 
lattice::densityplot(~bias,data=myDataMyPollsters,
            groups=pollster,
            xlab="Bias",
            main="Pollster Bias",
            plot.points=TRUE,
            auto.key=TRUE)
```




We can start this exploration by looking into bias modeled as response variable to year, categorized by pollster. It should be noted that this data excludes polls from special elections. A plot of bias by pollsters is plotted as dependent on time. 

```{r}
#plot by year and pollster
library(ggplot2)
ggplot(myDataMyPollsters, aes(x=year, y=bias, color=pollster)) +
  geom_point() +
  geom_smooth(method=lm)
```

We can also take a look at the relationship between year, bias, and election type:

```{r}
ggplot(myDataMyPollsters, aes(x=year, y=bias, color=type_simple)) +
  geom_point() +
  geom_smooth(method=lm)
```

Of interest is also the relationship between partisan classification and the bias of poll results. A plot is exploring this is shown below:

```{r}
ggplot(myDataMyPollsters, aes(x=year, y=bias, color=partisan)) +
  geom_point() +
  geom_smooth(method=lm)
```

As seen above, there is overwhelmingly little partisan polls considered in our data, making it hard to make any credible statements about their apparent effect on bias. Because of this, the partisan status of polls will not be used to model polling bias.

Another factor worth considering is the sample size of our polls. Because there are a few very large polls, we will perform a log transform on the numeric sample size variable. The bias plotted as a response variable to the log-transformed sample size variable, color coded by pollster can be seen below:

```{r}
datalog = transform(myDataMyPollsters, logsamplesize = log(samplesize))

ggplot(datalog, aes(x=logsamplesize, y=bias, color=pollster)) +
  geom_point() +
  geom_smooth(method=lm)


```

The model considered is seen below:
$$ \mu  [bias \~ pollster, year, type, samplesize] $$
```{r, include=FALSE}
logmodel = lm(bias ~ year*pollster + year*type_simple + year + type_simple + pollster + logsamplesize , data = datalog)

library(gtsummary)
library(dplyr)
```

A summary for the coefficients of this model can be seen below:

```{r}

tbl_regression(logmodel)

squared=summary(logmodel)$adj.r.squared
```

### Sampling Variablity Assumptions

To make sure that we can use model, we must check some sampling variability Assumptions.

Q-Q plot seems to trail off at the tails, but it appears about 95% of the points are within the 95% bounds.

The residual density plot seems centered on zero and mostly normal, with some outliers. It seems that this distrubtion matches a t-distrubtion fairly wel

The Cramer V value is only .203, indicating acceptable levels of correlation between our two discrete explanatory variables of type of poll and the pollster.

```{r}
#SVAs
car::qqp(rstandard(logmodel))
lattice::densityplot(rstandard(logmodel))
# the residual density plot looks cleary centered on zero and mostly normal with outliers
# robust linear modeling will probably do a very good job here when we construct our Bayesian model

#multicollinearity?
#Calculate Cramer's V,  a measure of intercorrelation of two discrete variables
tab <- xtabs(~pollster + type_simple, data = myDataMyPollsters)
corma=vcd::assocstats(tab)
summary(corma)
#20.3% isn't bad between our explantory variables

```

### Conclusions

While this model certainly isn't perfect, it has an adjusted R-squared value of only `r squared`, it can be helpful in helping us begin to understand how to model polling data. The final terms included in this model that are included as predictors are shown below:

* year:pollster
* year:type_simple
* year
* pollster
* type_simple
* log(samplesize)

\newpage
## Works Cited

---
#.bib for bib refrence
nocite: |
  @538rawpolldata
---
