---
title: "Comparing Pollsters"
author: "Haley Reed"
date: "6/5/2020"
output: html_document
bibliography: electionworkingbib.bib
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = dirname(getwd()))
```



While looking at one pollster may provide us with some relevant information, it is much more interesting to look at multiple pollsters and compare them. Choose your group of pollsters below (and dates for which you wish to see a comparison):

```{r, echo=TRUE}
pollsters= c("Rasmussen Reports/Pulse Opinion Research","Monmouth University", "Marist College", "ABC News/The Washington Post","SurveyUSA")

#Establish which years you wish to use data from (between 1998 and 2020):
earlyYear=1998
lateYear=2020
```

This file uses data from the FiveThirtyEight's repository to isolate the specified pollsters and their most recent polls at the time of every election in the dataset for the indicated years.

We can see the bias from the last poll for every election from each pollster. A segment of that data is shown in the table below:
```{r, echo=FALSE}

#load the data
#in this case from 538
mydata <- read.csv("raw-polls_538.csv")
```
  
```{r}



#use lubridate to change dates from character to date data type for functionality. 
mydata$electiondate = lubridate::mdy(mydata$electiondate)
mydata$polldate = lubridate::mdy(mydata$polldate)

#Subset for year constraints given above.
mydata=(subset(mydata, mydata$year>=earlyYear))
mydata=(subset(mydata, mydata$year<=lateYear))



#create empty data frame, while maintaining all columns from the mydata structure
myDataMyPollsters=  mydata[0,]

#run a loop to fill this data frame with every pollster specified above
for(myPollster in unique(pollsters))
{
  #subset to a dataset with just each pollster
  subpoll=subset(mydata, pollster==myPollster)
  #Use only the most recent poll for each election:
  #Use setDT function from data.table package to get a subset from mydata with just the max. value of the date element for each race (grouped with the keyby function). Call this new subset of data onlyRecentData.
  onlyRecentData1=(data.table::setDT(subpoll)[,.SD[which.max(polldate)],keyby=race_id])
  #combine each of these subsets together
  myDataMyPollsters=rbind(myDataMyPollsters, onlyRecentData1)
}



#create new data frame.
df1 <- data.frame(matrix(ncol = 7, nrow = 0))
#First two columns will just be Race and Election Date
x <- c("Race","Election Date")


#define x as the column names for our new dataframe.
colnames(df1) <- x

#loop through pollster filtered data to fill the data frame
for(myRace in unique(myDataMyPollsters$race_id))
{
  #subset the data to only one race at a time
  subrace=subset(myDataMyPollsters, race_id==myRace)
  #Choose a year by picking year of first data point in the frame
  myYear=subrace$year[1]
  RaceName=subrace$race
  
  #create new data frame to store info from each race
  newline=data.frame("Race"=RaceName, "Election Date"=myYear)
  
  #Loop through each pollster to find their bias
  for(myPollster in unique(pollsters))
  {
    #get bias for that pollster
    pbias=subset(subrace, pollster==myPollster)$bias
    if(length(pbias)==0){
      pbias=NA
    }
    #add that pollster's bias to individual race's data frame
    newline = cbind(newline, pbias)
  }
  
  
  #add the individual race data frames together 
  df1=rbind(df1, newline)
}

#loop through list of desired pollsters for the rest of the columns
for(myPollster in unique(pollsters))
{
  #Append pollster names to column name list
  x=c(x,myPollster)
}

#define x as the column names for our new dataframe.
colnames(df1) <- x


#show a table in RMarkdown
knitr::kable(head(df1))
```

Since there are concerns about the effect of herding on polling credibility, it is important to look at the correlation matrix shown below:

```{r}
#scatterplot matrix to show correlation
lattice::splom(subset(df1, select=pollsters))

```

Additionally, a density plot that compares pollster error can be seen below:

```{r}
#density plot 
lattice::densityplot(~bias,data=myDataMyPollsters,
            groups=pollster,
            xlab="Bias",
            main="Pollster Bias",
            plot.points=TRUE,
            auto.key=TRUE)
```

\newpage
## Works Cited

---
#.bib for bib refrence
nocite: |
  @538rawpolldata
---
