---
title: "MarginVersusPartyError"
author: "Haley Reed"
date: "7/3/2020"
output: html_document
bibliography: electionworkingbib.bib
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, error=FALSE, warning=FALSE)
knitr::opts_knit$set(root.dir = dirname(getwd()))
```

**Introduction**

In looking at past polls and elections, the "bias" of a poll is defined around the margin of difference between the two primary candidates of a race. 

What is the relationship between that "bias" and the actual number of percentage points a particular poll is off in judging candidates in a vacuum? 

What is the relationship between Republican Error, Democratic error, and traditional bias with the following definitions?

$$DemocraticError = DemocraticCanditatePollingPercent-DemocraticActualPercentage$$

$$RepublicanError = RepublicanCanditatePollingPercent-RepublicanActualPercentage$$

$$Bias = MarginPoll - MarginActual$$

$$MarginPoll = DemocraticCanditatePollingPercent-RepublicanCanditatePollingPercent$$
$$MarginActual = DemocraticActualPercentage-RepublicanActualPercentage$$


```{r, include=FALSE, fig.show='hide'}
source(here::here('RScripts', 'Exploration of Margin v. Party Error.r'))
```

Democratic and Republican year leans are shown below for comparison. As well as these year leans plotted together...
```{r}
knitr::kable(dtD)
knitr::kable(dtR)

plot(dtD$yearLabels,dtD$Mode, col="blue", main="Year Leans", xlab="Election Year", 
     ylab="Most Credible Error Value", ylim=c(-5, 1) , xlim=c(1996, 2020), xaxt = "n"
     )
# Define the position of tick marks
axis(side = 1, 
     at = yearLabels, 
     labels = yearLabels,
     tck=-.05)
points(dtR$yearLabels, dtR$Mode, col="red")
# Add a legend
legend(1996, 1, legend=c("Democratic Error", "Republican Error"),
       col=c("blue", "red"), lty=1:2, cex=0.8)
```




For matching points by year and pollster, the relationship between Democratic error and republican error does not look very strong....

In fact the R squared values is less than 0.


```{r}
plot(dtDpollster$Mode,dtRpollster$Mode, main="Democratic Versus Rebulican Pollster Biases, 2000-2016", xlab="Democratic Pollster Bias", 
     ylab="Republican Pollster Bias")
```

**Conclusion**

The relationships between Republican Error, Democratic error, and traditional bias are not as strong as we might initally think. 

\newpage

**Works Cited**

---
#.bib for bib refrence
nocite: |
  @538rawpolldata
  @KrusckeTextbook
---