---
title: "Pollster Analyzer"
author: "Haley Reed"
date: "5/29/2020"
output: html_document
bibliography: electionworkingbib.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = dirname(getwd()))
```

This file uses data from the FiveThirtyEight repository to isolate an individual pollster and their most recent polls at the time of every election in the dataset for the indicated years.

```{r, echo=FALSE}
#load the data
#in this case from 538
mydata <- read.csv("Data/raw-polls_538.csv")
```

```{r, echo=TRUE}
#Establish which pollster you wish to analyze:
pollsterOfChoice="Mitchell Research & Communications"
#Establish which years you wish to use data from (between 1998 and 2020):
earlyYear=1998
lateYear=2020
```

This file will analyze data from `r pollsterOfChoice` between the years of `r earlyYear` and `r lateYear`. This particular subset of data is shown in the table below:

```{r, include=FALSE}
#Subset for constraints given above.
mydata=(subset(mydata,pollster==pollsterOfChoice))
mydata=(subset(mydata, mydata$year>=earlyYear))
mydata=(subset(mydata, mydata$year<=lateYear))

#use lubridate to change dates from character to date data type for functionality. 
mydata$electiondate = lubridate::mdy(mydata$electiondate)
mydata$polldate = lubridate::mdy(mydata$polldate)

```

```{r, echo=FALSE}
#Use only the most recent poll for each election:
#Use setDT function from data.table package to get a subset from mydata with just the max. value of the date element for each race (grouped with the keyby function). Call this new subset of data onlyRecentData.
onlyRecentData=(data.table::setDT(mydata)[,.SD[which.max(polldate)],keyby=race_id])
#Create a table for onlyRecentData with knitr package.
knitr::kable(head(onlyRecentData))
```

The plot and summary statistics for margin of error for `r pollsterOfChoice` between the years of `r earlyYear` and `r lateYear` is shown below.

```{r, echo=FALSE}
#plot error in polling (absolute value of bias) as a dependent variable of the date of the election. 
plot(onlyRecentData$electiondate,onlyRecentData$error, main="Margin of Error of Polls", xlab="Election Date", ylab="Margin of Error")
#create a base 0 line - visualizes where a poll with no error would be.
abline(0,0,col="blue")  

#provide summary statistics for the error.
summary(onlyRecentData$error)
```

The plot and summary statistics for bias towards Republican candidates is shown below:

```{r, echo=FALSE}
#To check for partisan bias, will subset data down to just the races where candidate 1 and candidate 2 are running on two different party tickets.
twoPartyTicket=subset(onlyRecentData,cand1_party!=cand2_party)
#defensive programming against NA values for bias (appears when one of the candidates is an independent) 
twoPartyTicket=subset(twoPartyTicket, !is.na(bias))
#plot bias towards republicans in polling as a dependent variable of the date of the election.
plot(twoPartyTicket$electiondate,twoPartyTicket$bias, main="Bias of Polls towards Republican Candidates", xlab="Election Date", ylab="Bias")
#create a base 0 line - visualizes where a poll with no bias would be. 
abline(0,0,col="blue")
#provide summary statistics for the bias.
summary(twoPartyTicket$bias)
```


The density plot for the bias that `r pollsterOfChoice` shows for republican candidates between the years `r earlyYear` and `r lateYear` is shown below.
```{r}
#use density function of bias in the twoPartyTicket dataset to plot.
plot(density(twoPartyTicket$bias), main="Density Plot of Bias", xlab="Rebulican Bias")
```

Additionally, `r pollsterOfChoice` correctly predicted `r mean(onlyRecentData$rightcall)*100`% of races in their last poll for every election polled from `r earlyYear` to `r lateYear`.

\newpage
### All Polls

After exploring the data with only the most recent polls before every election, we can compare these findings to data that includes all polls. What is the difference between their error? Their bias? Their predictive power?

The plot and summary statistics for margin of error from all polls conducted by `r pollsterOfChoice` between `r earlyYear` and `r lateYear` is shown below.

```{r, echo=FALSE}
#plot error in polling (absolute value of bias) as a dependent variable of the date of the election. 
plot(mydata$electiondate,mydata$error, main="Margin of Error for all Polls", xlab="Election Date", ylab="Margin of Error")
#create a base 0 line - visualizes where a poll with no error would be.
abline(0,0,col="blue")  

#provide summary statistics for the error.
summary(mydata$error)
```

The plot and summary statistics for bias towards republican candidates for all polls conducted by `r pollsterOfChoice` between `r earlyYear` and `r lateYear` is shown below.

```{r, echo=FALSE}
#To check for partisan bias, will subset data down to just the races where candidate 1 and candidate 2 are running on two different party tickets.
alltwoPartyTicket=subset(mydata,cand1_party!=cand2_party)
#defensive programming against NA values for bias (appears when one of the candidates is an independent) 
alltwoPartyTicket=subset(alltwoPartyTicket, !is.na(bias))
#plot bias towards republicans in polling as a dependent variable of the date of the election.
plot(alltwoPartyTicket$electiondate,alltwoPartyTicket$bias, main="Bias of Polls towards Republican Candidates for all Polls", xlab="Election Date", ylab="Bias")
#create a base 0 line - visualizes where a poll with no bias would be. 
abline(0,0,col="blue")
#provide summary statistics for the bias.
summary(alltwoPartyTicket$bias)


#evaluate if including all polls or just the most recent ones did a better job.
evaluatePredictive=mean(onlyRecentData$rightcall)>mean(mydata$rightcall)
#evaluate the predictive difference.
predictiveDiff=mean(onlyRecentData$rightcall)-mean(mydata$rightcall)
```


The mean bias for every year is displayed in the table and graph below.  

```{r}
#To explore whether the bias of a pollster depends on the year...
#create new data frame to easily see average bias each year
Year = alltwoPartyTicket$year
Bias = alltwoPartyTicket$bias
df <- data.frame(Year, Bias)

#aggregate the data to find the mean for each year
yearmeans=aggregate(.~Year, data=df, mean)
#display this table of mean biases for each year in r. 
knitr::kable(yearmeans)

#plot averaged yearly bias as a dependent variable of time in years
plot(yearmeans$Year, yearmeans$Bias, main="Averaged Yearly Bias", xlab="Year", ylab="Bias towards Republican Party")
#create a base 0 line - visualizes where a poll with no bias would be. 
abline(0,0,col="blue")
```



For `r pollsterOfChoice` the standard deviation of averaged yearly biases is `r sd(yearmeans$Bias)`. 


Additionally, `r if(evaluatePredictive==TRUE){"using just the most recent poll from each election predicted more races correctly than using all polls."}` `r if(evaluatePredictive==FALSE){"using all the polls from every election predicted more races correctly than using just the most recent poll from each election."}`
`r pollsterOfChoice` correctly predicted `r mean(mydata$rightcall)*100`% of races using all polls from every election polled from `r earlyYear` to `r lateYear`.

\newpage
## Works Cited

---
#.bib for bib refrence
nocite: |
  @538rawpolldata
---
